Reducing RES Droughts through the integration of wind and Solar PV

### 2025-03-28 Article/RES_droughts_Morin_ReEn_R2.tex
Modify seasonality and conclusions

The seasonal variations observed in this study have important implications for energy planning. Given that energy demand peaks in winter for Northern European countries, understanding these seasonal patterns is critical for assessing the need for conventional backup or storage solutions during periods of prolonged low renewable output. The findings underscore that even small differences in model assumptions leads to significant variations in drought estimates, thereby affecting the reliability of the energy system during critical periods. Additionally, the integration of large shares of solar PV in the system leads to a general reduction of RES droughts year-round. However, it can not counter the presence of RES droughts in the winter months due to its natural limitations, inevitably leading to higher reserve capacity needs during winter months as reliance on RES increases. Such insights are essential for policymakers to develop targeted strategies that enhance grid resilience and ensure a stable energy supply throughout the year.

@@

This study has explored the characterisation of RES droughts in the real-life transition from a wind-dominated system to a more balanced system with integrated solar PV. This has been done through the comparison over a 45-year period of three different datasets: one based on a hand-crafted validated model and two based on a generic model, C3S-Energy. The generic model has two versions, one using large-scale aggregated information, and one which includes the locations of farms as well (also considered in the validated model).

Our results show that generic datasets present limitations in the quantification of RES droughts. Although the three datasets used capture overall trends in drought occurrence, significant differences emerge when using non-tailored data for the study of RES droughts. This finding highlights that the choice of dataset and its underlying assumptions can lead to consequential differences in estimated RES drought characteristics, emphasising the need for datasets that are specifically designed for extreme event analysis.

This study reveals that differences in model parametrisation, particularly in the representation of wind turbine power curves and solar PV panel characteristics, have a stronger influence on the drought estimates than the inclusion of RES farm locations. The use of a validated dataset with a carefully selected wind turbine power curve consistently produced higher return periods and a greater number of drought events for wind energy than the datasets derived from C3S-Energy. This suggests that fine-tuning model parameters to match observed data is crucial for accurately quantifying RES drought risks, thereby supporting more effective energy system planning.

Finally, the effect on RES droughts of the integration of solar PV in a wind-dominated system has been explored in a real-case setting. In the presented example of Ireland, the analysis has demonstrated that transitioning to a more balanced system with similar amounts of wind and solar PV markedly reduces the frequency, duration, and seasonal variability of drought events. This improvement is attributed to the complementary nature of wind and solar PV generation, as solar PV typically peaks in summer while wind generation is more consistent in winter. Thus, a more diversified renewable energy mix not only mitigates extreme drought conditions but also enhances overall system resilience, providing valuable insights for policymakers tasked with ensuring energy security.

The results presented in this study have several limitations. Although ERA5 is among the best reanalysis datasets for renewable energy analysis, it still presents some biases, and its resolution may not capture local-scale phenomena. This is especially limiting if individual farms are considered, but its still shows good skill at statistically representing nation-wide behaviours considering the distribution of farms. Moreover, the methodology employs a fixed threshold to define RES drought events, which is necessary for comparing the three datasets considering only weather-derived drivers, but does not account for demand variations. Consequently, while this approach enables a consistent inter-comparison, it will not signal events that are most critical for power system operations for their mismatch of RES generation and demand.

>>>>

The seasonal variations of RES droughts observed in this study have important implications for energy planning. Energy demand peaks in winter for Northern European countries, making the seasonality of RES droughts critical for the sizing of reserve capacity. Our results show that selecting the wrong dataset could severely underestimate RES droughts during winter months, thereby affecting the reliability of the energy system during critical periods. Additionally, the integration of large shares of solar PV in the system leads to a generalised reduction of RES droughts, yet winter months present a slight increase. The natural limitations of solar PV lead to inevitably higher reserve capacity needs during winter months as reliance on RES increases. These types of insights are essential to develop targeted strategies that enhance grid resilience and ensure a stable energy supply throughout the year.

@@

This study has explored the characterisation of RES droughts in the transition from a wind-dominated system to a more balanced system with integrated solar PV, based on the real case of Ireland. Three different datasets were compared over a 45-year period: one created using a regionally validated model and two derived from a generic dataset for Europe, C3S-Energy. The two datasets derived from C3S-Energy present different approaches, with one using large-scale aggregated information only, and the other one including the locations of farms as well. The regionally validated model considered the locations of farms as well as tailored wind and PV models selected to best represent the actual generation in Ireland.

Our results show the limitations in the quantification of RES droughts present in datasets that have not undergone regional validation. The three datasets used in this study are able to capture overall trends in drought occurrence such as the seasonal cycle or the effect of increasing the share of solar PV. However, significant differences in the quantitative values, particularly the extremes, emerge when using non-validated datasets for the study of RES droughts. This finding highlights that using a non-validated dataset can lead to undersized reserve capacity, with the associated negative consequences for grid stability and security of supply.

This study has also revealed that differences in the wind turbine power curves and solar PV panel models have a stronger influence on the estimation of RES droughts than the consideration of RES farm locations. The two datasets derived from C3S-Energy consistently underestimated the number of wind drought events and the frequency of extremes when compared to the regionally validated dataset with a specifically selected wind turbine power curve. This suggests that a meticulous selection of the wind turbine power curve to match observed data is crucial for accurately quantifying RES drought risks, thereby supporting more effective energy system planning.

Finally, the effect of the integration of solar PV in a wind-dominated system on RES droughts has been explored in a real-case setting based on Ireland. Our analysis has demonstrated that transitioning to a system with similar amounts of wind and solar PV reduces the frequency, duration, and seasonal variability of drought events. This improvement is attributed to the complementary nature of wind and solar PV generation, as solar PV typically peaks in summer while wind generation is more consistent in winter. However, this integration is unable to counter the critical winter RES droughts, which coincide with the strongest electricity demand in Northern European countries like Ireland. Still, a more diversified renewable energy mix mitigates extreme drought conditions and enhances overall system resilience.

The results presented in this study have three main limitations. First, the source for weather data is ERA5, which is among the best reanalysis datasets for renewable energy applications, but still presents some biases and limited spatial resolution. Second, the presented study uses a fixed threshold to define RES drought events, but other methods could yield different results, even though we would expect the main takeaways to be the same. Third, the definition of RES droughts based on generation does not consider the mismatch between renewable generation and demand, which may be of interest to system operators, as these events put large amounts of strain on the system.
